---
title: "G5058 Assignment 1 Answer Key"
author: "Ben Goodrich"
date: "September 25, 2018"
output: html_document
---

```{r include=FALSE}
set.seed(12345)
```


# 1. Data Structures 

This problem was created by executing
```{r eval=FALSE}
types <- c("vector", "matrix", "array", "list", "data.frame")
sample(types, size = 2, replace = FALSE)
```
and then asking how these two types differ and how each can be created.

As discussed in lecture and in chapter 2 of Wickham, the following table

?d | Homogenous       | Heterogeneous
-- | ---------------- | -------------
1d | `vector`         | `list`
2d | `matrix`         | `data.frame`
nd | `array`          |     

illustrates the main differences between any two of these types. Either
the differ by dimensionality or whether their elements are homogenous.

We can use the `c` function to create a `vector`
```{r}
num <- c(3.5, 10)           # intermediate flexibility
int <- c(-2L, 4L)           # intermediate flexibility
char <- c("hello", "world") # most flexible
logic <- c(TRUE, FALSE)     # least flexible
```
the `matrix` function to create a matrix
```{r}
mat <- matrix(1:6, nrow = 3, ncol = 2)
mat
```
the `array` function to create an array 
```{r}
arr <- array(1:24, dim = c(4, 2, 3)) # 4 rows, 2 columns, 3 shelves
arr
```
the `list` function to create a list
```{r}
lst <- list(logic, mat, arr)
```
and the `data.frame` function to create a data.frame
```{r}
df <- data.frame(num, char, logic)
```

# 2. Style

Here is some poorly styled R code:
```{r eval = FALSE}
a < - c( 5,-2,3,-4,1,NA )
  b < - a * - 1
  b[ b>0 ]

num = 1
mycondition <- round (runif (1,0,1,))
if( mycondition ) {
    num <- num + 1
    print(paste("num =" , num , sep=" "))
} else {
print("false")
}

# make an integer sequence from 1 to 50
y <- 1:50

# test if an number is even
even <- y%%2 = = 0

# subset by the test above
y = y [even]

x <- matrix( c( 23, 34, 35, 6, 87, 39, 21, 14, 99 ) ,nrow = 3)
df <- as.data.frame (x)
names( df ) <- c("percentage_score_on_reading_test", 
 "percentage.score.on.math.test", 
  "percentage-score-on-writing-test")

mean <- function (x)
{
sum(x)/length(x)
}
```
We revise it to have good formatting
```{r eval = FALSE}
# revised with proper spacing
a <- c(5, -2, 3, -4, 1, NA)
b <- a * -1
b[b > 0]

# revised assignment, indentation, and spacing
num <- 1
my_condition <- round(runif(1, 0, 1))
if (my_condition) {
  num <- num + 1
  print(paste("num =", num, sep=" "))
} else {
  print("false")
}

# eliminate useless comments, use proper spacing
y <- 1:50
even <- y %% 2 == 0
y <- y[even]

# shorter names, proper spacing
x <- matrix(c(23, 34, 35, 6, 87, 39, 21, 14, 99), nrow = 3)
df <- as.data.frame(x)
names(df) <- c("perc_read", "perc_math", "perc_write")

# better indentation and spacing plus correct brace placement
# best to not give a function the same name as an existing function
simple_mean <- function(x) {
  return(sum(x) / length(x))
}
```
Wickham does not always use the `return` statement, but it is more 
explicit to utilize it.

# 3. Subsetting a data.frame

This problem asked you to take a randomly selected state and two randomly selected variables via
```{r}
(state <- sample(rownames(state.x77), 1))
(vars <- sample(colnames(state.x77), 2, replace = FALSE))
```
and extrat the data on these variables for that state in three different ways.

1. By name:
```{r}
state.x77["Oregon", c("Area", "HS Grad"), drop = FALSE]
```
2. By index:
```{r}
state.x77[37, c(8, 6), drop = FALSE]
```
3. By logical vector:
```{r}
state.x77[rownames(state.x77) == "Oregon", 
          colnames(state.x77) %in% c("Area", "HS Grad"), drop = FALSE]
```
4. By **dplyr** syntax:
```{r, message = FALSE}
library(dplyr)
select(as.data.frame(state.x77), c("Area", "HS Grad")) %>%
  filter(., rownames(.) == "Oregon")
```

# 4. Manipulating data

This question asked you to perform one of five randomly selected tasks involving
the `data.frame` called `airquality`:
```{r}
str(airquality)
```

1)
Drop the last six rows from the data.frame. 
```{r}
airquality2 <- airquality[-c((nrow(airquality) - 5):nrow(airquality)),]
```

2)
Replace the `Month` variable with the month represented as an ordered factor.
```{r}
airquality$Month <- factor(airquality$Month, levels = 1:12, 
                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                      "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"), ordered = TRUE)
table(airquality$Month)
```

3)
Replace the `Wind` variable with windspeed measured in kilometers per hour.
```{r}
airquality$Wind <- airquality$Wind * 1.609344 # now in km / hour
head(airquality)
```

4)
Replace the missing values on the Solar.R variable with the average value of Solar.R among non-missing values.
```{r}
airquality$Solar.R[is.na(airquality$Solar.R)] <- 
  mean(airquality$Solar.R, na.rm = TRUE)
head(airquality)
```

5)
Replace the Temp variable with temperature measured in Celsius.
```{r}
airquality$Temp <- (airquality$Temp - 32) * (5/9) # now in Celsius
head(airquality)
```

# 5. American National Election Study

These questions were randomly assigned from Kosuke Imai's 
[book](Quantitative Social Science: An Introduction), specifically
Excercise 1.5.1 in [Chapter 1](http://assets.press.princeton.edu/chapters/s11025.pdf)

```{r}
data("turnout", package = "qss")
```

> Calculate the turnout rate based on the voting age population or VAP. Note that
for this data set, we must add the total number of eligible overseas voters since
the VAP variable does not include these individuals in the count. Next, calculate
the turnout rate using the voting eligible population or VEP. What difference do
you observe?

```{r}
round(with(turnout, cbind(year = year, 
                          VAP = total / (VAP + overseas),
                          VEP = total / (VEP + overseas))), digits = 2)
```
The main difference is that VEP > VAP

> Compute the difference between the ANES estimate of turnout rate and the
ratio of total to VAP across years. How big is the difference on average?
What is the range of the differences?

```{r}
statistics <- with(turnout, cbind(year = year, 
                                  VAP = total / VAP, 
                                  VEP = total / VEP,
                                  ANES = ANES / 100,
                                  diff = total / VAP - ANES / 100))
mean(statistics[ , "diff"])
range(statistics[ , "diff"])
```

> Compare the ANES turnout rate to the ratio of total to VEP for the presidential elections 
in 1980, 1984, 1988, 1992, 1996, 2000, 2004, and 2008 with those in midterm elections in
1982, 1986, 1990, 1994, 1998, and 2002 (2006 is omitted). Does the bias of the ANES 
estimates vary across election types?

```{r}
presidential <- c(1980, 1984, 1988, 1992, 1996, 2000, 2004, 2008)
```

```{r}
round(statistics[statistics[ , "year"] %in% presidential, ], digits = 2)
round(statistics[!(statistics[ , "year"] %in% presidential), ], digits = 2)
```
In general, the difference between `VEP` and `ANES` is larger during
presidential election years, although the midterm election in $2002$
was an exception.

> ANES does not interview prisoners and overseas voters. Calculate an adjustment
to the 2008 turnout rate. Begin by subtracting the total number of ineligible
felons and noncitizens from the VAP to calculate an adjusted VAP. Next,
calculate an adjusted VAP turnout rate, taking care to subtract the number of
overseas ballots counted from the total ballots in 2008. Compare the adjusted
VAP turnout with the unadjusted VAP, VEP, and the ANES turnout rate. Briefly
discuss the results.

```{r}
round(with(turnout, cbind(year = year,
                          VAP_ = VAP - felons - noncit,
                          turnout_ = (total - ifelse(is.na(osvoters), 0, osvoters)) / 
                                     (VAP - felons - noncit),
                          ANES = ANES / 100)), digits = 2)
```
The results are in the same direction as before, but the adjustments do make
a noticeable difference.
