---
title: "G5058 Week 09"
author: "Ben Goodrich"
date: "October 30, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval = FALSE}
install.packages(c("lars", "caret", "elasticnet", "mlbench")) # do this once outside
```

# Least Squares Estimator

Suppose that you have a $N \times p$ matrix of predictors, $\mathbf{X}$, that 
typically includes a column of ones, and an outcome vector, $\mathbf{y}$. You
think that the outcome is generated as a linear function of the predictors
$$y_i = \sum_{k=1}^p x_{ik} \boldsymbol{\beta}_k + \epsilon_i$$
with weights $\boldsymbol{\beta}$ and errors $\boldsymbol{\epsilon}$.

The least squares estimator is 
$\left(\mathbf{X}^\top \mathbf{X}\right)^{-1} \mathbf{X}^\top \mathbf{y}$
provided that $\left(\mathbf{X}^\top \mathbf{X}\right)^{-1}$ exists, which
it typically will if $N > p$ unless some column of $\mathbf{X}$ is an
_exact_ linear function of the other columns in $\mathbf{X}$. If you are an 
econometrician or statistician who is interested in making an inference
about $\boldsymbol{\beta}$ in the population from a finite but random sample 
of size $N$ from that population, then the absence of a _unique_ solution 
for the coefficients is a devestating problem. If you a data miner who is
only interested in making predictions for the outcomes, then the lack of
a unique solution is no big deal because _any_ solution yields the same
predictions $\widehat{\mathbf{y}} = \mathbf{X} \mathbf{b}$. However, you
might and should be worried that the error when predicting _this_ 
$\mathbf{y}$ is less than the prediction error when predicting _future_
outcomes.

# Training and Testing

In supervised learning, $\mathbf{X}$ and $\mathbf{y}$ are called the "training data",
which are used to obtain $\mathbf{b}$ in this case or more generally to solve
some optimization problem. Let $\mathbf{X}^\ast$ and $\mathbf{y}^\ast$ denote
the corresponding matrix of predictors and outcome vector in the "testing data",
which are additional observations that are _not_ used to obtain $\mathbf{b}$
and are instead used to evaluate the model that produced $\mathbf{b}$.

For supervised learning, the only question that matters is how well does the
model predict $\mathbf{y}^\ast$, which can be evaluated by several criteria,
such as 
$\frac{1}{N^\ast} \mathbf{e}^\ast\left(\mathbf{b}\right)^\top \mathbf{e}^\ast\left(\mathbf{b}\right)$, where $N^\ast$ is the number of 
testing observations and
$$\mathbf{e}^\ast\left(\mathbf{b}\right) = \mathbf{y}^\ast - \mathbf{X}^\ast \mathbf{b}.$$

For example,
```{r}
prostate <- read.table("http://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data",
                       header = TRUE)
str(prostate, max.level = 1)
prostate[,1:8] <- scale(prostate[,1:8]) # data miners usually standardize
```
The `prostate` dataset already has a logical vector called `train` that was
created somehow (often randomly).
```{r}
library(dplyr)
training <- select(filter(prostate, train == 1), -train)
testing <- select(filter(prostate, train == 0), -train)
c(training = nrow(training), testing = nrow(testing))
```

# The `lm` function

The most convenient way to obtain $\mathbf{b}$ in R is via the `lm` function,
which you are probably already familiar with from your statistics classes.
```{r}
ols <- lm(lpsa ~ ., data = training)
```
where the `.` means include all variables in `training` that are not otherwised
mentioned (like `lpsa` is)

However, in G5058, you should not utilize the output of `lm` for 
_statistical_  purposes, such as calculating standard errors, test statistics, 
$p$-values, confidence intervals, etc. Those all presume that your data is a random 
sample from a population and that you are interested in the probability 
distribution of $\mathbf{b}$ over repeated samples of size $N$ from this 
population. In supervised learning, the data are almost never a random sample 
from a population and even if they were, data miners do not care about the 
probability distribution of $\mathbf{b}$ over repeated samples of size $N$ from 
that population. Just like you can lose points for working R code that does
not utilize proper style, you can lose points for referring to R output that
relies on statistical assumptions in order to make statistical inferences.

You may think that a coefficient estimate that is "statistically significant"
would also imply the variable is a good predictor of the outcome, but you
would sometimes be wrong. Tian Zheng in the statistics department at Columbia
co-authored a paper
http://www.pnas.org/content/112/45/13892.long
explaining "Why significant variables aren’t automatically good predictors",
which QMSS does not really give you the tools to understand fully. I would say
the essence of it is that predictions are a function of the _entire_ $\mathbf{X}$
matrix and whether a variable adds predictive ability depends on how it combines
with the other variables in $\mathbf{X}$. Whether a coefficient is statistically
significant adds additional assumptions that may not hold but also refers to a
scenario in which all the other variables in $\mathbf{X}$ are held constant in
order to isolate the effect of one variable.

Data miners typically would not even look at the solution vector of coefficients,
although it is possible to do so
```{r}
round(coef(ols), digits = 2) # the b vector
```

We can verify that 
$\left(\mathbf{y} - \mathbf{X}\mathbf{b}\right)^\top \mathbf{X}$ is a zero 
(row) vector using the `residuals` and `model.matrix` functions, which 
extract the residuals and the $\mathbf{X}$ matrix used to obtain them from
a previously-fitted object:
```{r}
X <- model.matrix(ols)
round(t(residuals(ols)) %*% X, digits = 14) # zeros by construction
```
One of the worst mistakes that people make is to take the fact that this is a
zero vector as _evidence_ that a linear model is correct. This merely defines
the estimator and holds regardless of whether the outcome is generated as a
linear function of these predictors.

We can also calculate the Mean Squared Error (MSE) in the _testing_ data 
using the `predict` function and the `newdata` argument:
```{r}
y_hat <- predict(ols, newdata = testing)
mean((testing$lpsa - y_hat) ^ 2)
```

# Cross validation

You will also see data miners use what is known as $k$-fold cross-validation, 
where you split the original dataset into $k$ (often $k = 10$) mutually exclusive
and exhaustive (and often equally sized) subsets, treat $k - 1$ of these subsets
collectively as the "training" data to obtain $\mathbf{b}$ and use the excluded 
subset as the "testing" data to evaluate $\mathbf{e}^\ast\left(\mathbf{b}\right)$. 
That can be done $k$ possible ways and you can average the results over them.

The **caret** package makes this relatively easy to do

```{r, message = FALSE}
library(caret)
ctrl <- trainControl(method = "cv", number = 10)
set.seed(12345)
ols_cv <- train(lpsa ~ ., data = training, method = "lm", trControl = ctrl)
ols_cv
```
In this particular case, the Root Mean Squared Error (RMSE) when predicting
the outcome in the held-out fold is actually _less_ than the estimated standard 
deviation of the error in the original model that used all the training data
```{r}
sigma(ols)
```
but usually the RMSE will be larger.

# Boston Housing Example

Step 0: Load the whole dataset
```{r}
data(BostonHousing, package = "mlbench")
```

Step 1: Split into training and testing
```{r, message = FALSE}
in_train <- createDataPartition(y = BostonHousing$medv,
                                p = 3 / 4, list = FALSE)
str(in_train)
BH_training <- BostonHousing[ in_train, ]
BH_testing  <- BostonHousing[-in_train, ]
```

Step 2: Solve an optimization problem using the pre-processed training data
```{r}
fit <- train(medv ~ ., data = BH_training, method = "lm", 
             preProcess = c("center", "scale"))
```

Step 3: Predict in the testing data
```{r}
y_hat <- predict(fit, newdata = BH_testing)
```

Step 4: Evaluate
```{r}
defaultSummary(data.frame(obs = BH_testing$medv, pred = y_hat))
```

A similar set of steps can be used for many of the supervised learning
methods we will learn for the rest of the semester in GR5058. However,
in many cases there are additional tuning parameters we have to deal
with.

# Subset selection

Forward selection starts with a model that just has an intercept and sequentially adds
predictors. Backward selection starts with the full model and sequentially subtracts
predictors. The `step` function in R, by default, uses a combination of both directions
and ranks the models by the Aikaike Information Criterion (AIC), which is defined as $-2$
times the "log-likelihood", plus $2$ times the number of parameters estimated. The
log-likelihood in the regression case is a function of 
$\mathbf{e}\left(\mathbf{b}\right)$.
```{r}
c(AIC = AIC(ols), formula = -2 * logLik(ols) + 2 * (length(coef(ols)) + 1))
```
As can be seen, the AIC consists of the sum of a "misfit" term and a "complexity" term.

Using the AIC is preferable to using the adjusted R-squared or $p$-values for model
selection because the model with the lowest AIC is _expected_ to predict best outside
the training data. Data miners tend to prefer to obtain the model that _actually_
predicts best in the particular testing data that they have and tend to have a dim
opinion of tools invented by statisticians such as the AIC.

```{r}
ols_subset <- step(ols)
round(coef(ols_subset), digits = 2)
setdiff(names(coef(ols)), names(coef(ols_subset)))
```

# Why penalize?

OLS chooses the coefficient values in order to minimize the sum-of-squared residuals.
Doing so is an _unbiased_ estimator for the coefficients. What does unbiased mean in 
the statistical sense? It means that if we were to sample $N$ observations from this
population many, many times, each time executing the OLS estimators, the element-wise
average of $\{\mathbf{b}^{[1]}, \mathbf{b}^{[2]}, \dots\}$ would be equal to 
$\boldsymbol{\beta}$. Using an unbiased estimator for the coefficients presumes the true 
coefficients in the population are of interest, which they may not be if we are only 
interested in prediction of the outcome.

Under some strong assumptions, OLS is the Best Linear Unbiased Estimator (BLUE), 
meaning that among all linear estimators that are unbiased, OLS has the smallest
variance in the coefficient estimates from one sample of size $N$ to the next.
However, if you do not care about the distribution of the coefficient estimates from
one random sample of size $N$ to the next, then it is quite possible to obtain a 
non-linear or biased estimator that yields better predictions in the testing data.

If the vector of residuals is given by 
$e\left(\mathbf{b}\right) = \mathbf{y} - \mathbf{X} \mathbf{b}$
there are a variety of estimators of the parameters in a linear model that minimize
something other than the sum-of-squared residuals (SSR), such as

* Ridge regression, which minimizes the SSR subject to the constraint that 
$\mathbf{b}^\top \mathbf{b} \leq s$, which can be reformulated with a Lagrange
multiplier as
$$e\left(\mathbf{b}\right)^\top e\left(\mathbf{b}\right) + \lambda \sum_{k=1}^p b_k^2$$
* Lasso, which minimizes the SSR subjec to the constraint that $\sum_{k=1}^p\left|b_k\right| < s$
which can be reformulated with a Lagrange multiplier as
$$e\left(\mathbf{b}\right)^\top e\left(\mathbf{b}\right) + \lambda \sum_{k=1}^p \left|b_k\right|$$
* Least Angle Regression, which does not have an explicit objective function but is
  sort of a combination of forward search and lasso.
  
Ridge regression and Lasso regression can be generalized to an objective function that
minimizes the SSR plus $\lambda$ times the sum of absolute coefficients raised to some
positive power $\left(\tau\right)$. 
$$e\left(\mathbf{b}\right)^\top e\left(\mathbf{b}\right) + 
\lambda \sum_{k=1}^p \left|b_k\right|^\tau$$
This can be given a Bayesian justification (but that does not make the people who utilize 
it Bayesians).

Least squares has $p$ unknowns and $p$ orthogonality constraints so there is a unique
solution (unless $\left(\mathbf{X}^\top \mathbf{X}\right)^{-1}$ does not 
exist). When you introduce "tuning parameters" such as $\lambda$, then there are
more unknowns than there are constraints. But for any given value of $\lambda$, then
you can solve the optimization problem to obtain the coefficients. So, how do you
choose the value for $\lambda$? The traditional answer is with some form of cross-validation
to see which value of $\lambda$ yields the best predictions outside the data that
are used to obtain the coefficients.

First, we need a set of $\lambda$ values to search over.
```{r, message = FALSE}
enetGrid <- expand.grid(.lambda = seq(.05, 1, length = 10),
                        .fraction = seq(.05, 1, length = 10))
head(enetGrid)
```
Then, we use **caret** to find the best fraction of the 
training data to use and the best value of $\lambda
```{r}
set.seed(12345)
lasso <- train(lpsa ~ ., data = training, method = "enet", 
               trControl = ctrl, tuneGrid = enetGrid)
lasso$bestTune
```

Actually looking at the coefficients is somewhat convoluted (and not intended)
```{r}
lambda <- lasso$bestTune$lambda
b_lasso <- coef(predict(lasso$finalModel, type = "coefficients", 
                        s = lambda, mode = "penalty"))
b_lasso
```
and we can see dropped the `lcp` variable from the model by giving it a 
coefficient of zero.

The `predict` method for the `lasso` object yields predictions in the
testing data (for the best $\lambda$):
```{r}
y_hat <- predict(lasso, newdata = testing)
mean( (testing$lpsa - y_hat) ^ 2 )
```

We can also do least angle regression in a similar fashion:
```{r}
enetGrid$.lambda <- NULL # delete the .lambda column
LARS <- train(lpsa ~ ., data = training, method = "lars", 
              trControl = ctrl, tuneGrid = enetGrid)
LARS
```

# Derived Input Methods

* Principal Components Regression, which involves replacing
  correlated predictors with an orthogonal approximation of the space they span
  If this involves two components, then it would be the dimensions of the 
  `biplot` but it can involve more than two retained components.
* Partial Least Squares, which is similar to Principal Components Regression but
  does a matrix decomposition including the outcome as well, making it a
  more supervised learning approach.

Partial Least Squares (PLS) is better for a variety of reasons.

In both cases, the key question is how many components to include in the 
approximation.
```{r, message = FALSE, warning = FALSE}
PLS <- train(lpsa ~ ., data = training, method = "pls", 
             tuneLength = 20, trControl = ctrl)
PLS
```

```{r}
y_hat <- predict(PLS, newdata = testing)
mean( (testing$lpsa - y_hat) ^ 2 )
```

# Bakeoff with Apartment Prices

There are two `data.frames` on CourseWorks that you can access by executing
```{r eval=TRUE}
training <- readRDS(gzcon(url('https://courseworks.columbia.edu/x/pJdP39')))
testing  <- readRDS(gzcon(url('https://courseworks.columbia.edu/x/QnKLgY')))
```
that each contain $109$ randomly selected observations on apartments for purchase
in a Western European city in $2005$. The dependent variable is `totalprice`, which
is the purchase price of the apartment in Euros. The possible predictors are:

* `area` the number of square meters in the apartment
* `zone` an unordered factor indicating what neighborhood the apartment is in
* `category` an ordered factor indicating the condition of the apartment
* `age` number of years since the apartment was built
* `floor` the floor of the building where the apartment is located
* `rooms` the total number of rooms in the apartment
* `out` an ordered factor indicating what percentage of the apartment's exterior
  is exposed to the outside
* `conservation` an ordered factor indicating how well the apartment is conserved
* `toilets` a count
* `garage` a count, i.e. some apartments have two garages
* `elevator` a binary variable
* `streetcategory` an ordered factor that captures the quality of the street the
  apartment building is on
* `heating` an unordered factor indicating something about the presence or absence
  of (possibly central) heating for the apartment
* `storage` a count of the number of storage rooms for the apartment

Use the `training` data.frame to estimate models that are _linear in their parameters_ 
and select the "best" such model according to which has the smallest average squared 
error when you predict `totalprice` in the `testing` data. You will need the `predict` 
function to do this.

Let's start with a scatterplot of `totalprice` (in hundred thousand Euros) and 
`area` (in square meters) with additional features of the apartment indicated:
```{r, message = FALSE}
training$toilets <- as.factor(training$toilets)
training$storage <- as.factor(training$storage)
testing$toilets <- as.factor(testing$toilets)
testing$storage <- as.factor(testing$storage)
library(ggplot2)
ggplot(training) + 
  geom_point(aes(x = area, y = totalprice / 1000, 
                 col = rooms, shape = toilets, 
                 fill = storage)) + 
  xlab("Area (square meters)") + ylab("Price (in 1,000 Euros)")
```

As can be seen, there is not a whole lot of variation in `totalprice` to exaplin
after the `area` of the apartment is taken into account. In particular, it looks
as if the number of rooms is superfluous after adjusting for `area` and the 
number of `toilets` (which are counted as a room each). In other words, four-room
apartments are more expensive than two-room apartments but four-room apartments
are proportionally larger in terms of `area` and almost always seem to have two 
`toilets` and a `storage` area.

This intuition is confirmed by the results of a regression
(where the outcome is now in raw Euros)
```{r}
ols <- lm(totalprice ~ area + rooms + toilets + I(storage != 0), 
          data = training)
summary(ols)$adj.r.squared
```
This model has an (adjusted) $R^2$ that is greater than $0.7$. Nevertheless,
a data miner would tend to use computational tools to try to arrive at a 
reasonably parsimonious model that predicted well (in the testing sample),
rather than starting there on the basis of (possibly flawed) intuition about
how real estate prices work.

Thus, we start with a model that includes all available predictors, an
interaction between `elevator` and `floor` (because the presence of an
elevator in the building is valuable, but not so much if you are on the
first floor) and the square of `age` (in case there are non-linearities).
Note that the `.` simply indicates all variables in `trainging` not otherwise 
in the formula.
```{r}
ols <- lm(totalprice ~ . + elevator * floor + I(age ^ 2) - toilets - storage, 
          data = training)
y_hat_OLS <- predict(ols, newdata = testing)
```
The `step` function can then be used to rank variations on this model
by the AIC.
```{r}
ols_AIC <- step(ols, trace = FALSE)
```
Somewhat surprisingly, the `step` function only drops a few predictors,
namely:
```{r}
setdiff(names(coef(ols)), names(coef(ols_AIC)))
```

Get the preditions
```{r}
y_hat_AIC <- predict(ols_AIC, newdata = testing)
```

We could then try lasso:
```{r}
enetGrid <- expand.grid(.lambda = seq(.05, 1, length = 10),
                        .fraction = seq(.05, 1, length = 10))
lasso <- train(formula(ols), data = training, method = "enet", 
               trControl = ctrl, tuneGrid = enetGrid)
y_hat_lasso <- predict(lasso, newdata = testing)
```

Then PLS
```{r}
PLS <- train(formula(ols), data = training, method = "pls", 
             tuneLength = 20, trControl = ctrl)
y_hat_PLS <- predict(PLS, newdata = testing)
```

To summarize, 
```{r}
defaultSummary(data.frame(obs = testing$totalprice, pred = y_hat_OLS))
defaultSummary(data.frame(obs = testing$totalprice, pred = y_hat_AIC))
defaultSummary(data.frame(obs = testing$totalprice, pred = y_hat_lasso))
defaultSummary(data.frame(obs = testing$totalprice, pred = y_hat_PLS))
```
