---
title: "G5058 Week 03"
author: "Ben Goodrich"
date: "September 18, 2017"
output: html_document
---

# CampusWire

If you have not already done so, go to
https://campuswire.com/p/G8F62DF08
and create an account using passcode 0010. Public questions and answers to conceptual
questions count toward your the class participation portion of your grade. Through
two weeks, most students have contributed literally nothing. Also, read the other
threads before posting the same question.

# Accessing the Course Server

The QMSS has arranged for people enrolled in G5058 to access RStudio on an
ISERP server. To do so, first ensure that you have a _strong_ password on
your Columbia account
https://cuit.columbia.edu/strong-passwords
Moreover, it is advisible to enable Multifactor Authentication on all your
Columbia logins
https://cuit.columbia.edu/mfa

Then, visit
https://rstudio.qmss.columbia.edu
and enter your _Columbia_ email address, using your UNI rather than any
alias (involving your name) that you may have set up. After clicking on
the icon that says "Sign into Columbia", you will be redirected to a
Columbia webpage where you enter your UNI (i.e. without the `@columbia.edu`)
and your aforementioned strong password. At that point, you should see a
RStudio client within your webbrowser like you see when I present lecture
material.

Remember, to do the following once on the server:

1. Click on __Tools -> Global Options__
2. Uncheck the box that says "Restore .RData into workspace on startup"
3. Change the drop-down menu that says "Save workspace to .RData on exit" to Never
4. Click Apply
5. Click OK

This server is primarily intended to be a fallback in case the RStudio or
necessary R packages are not working correctly locally on your laptop.
However, it is also correctly configured to knit RMarkdown files to PDFs
if you need that, rather than HTML. You may use this server for G5058 or
any other course, including the QMSS thesis course. You may not use this
server for "research", including RA work.

If you need to upload files from your laptop to the server, use the **Upload**
button under the **Files** tab in the bottom right plane. Conversely, if you
need to download files from the server to your laptop, click on the checkbox
to the left of the file you need, then click on the **More** drop-down menu 
and then click **Export**.

# Installing Packages

If your laptop (or less likely the server) is missing a necessary R package, 
you can install it (and their dependencies) by executing, for example,

```{r, eval = FALSE}
install.packages("beepr")
```
in the Console.

This only needs to be done once per computer and calls to `install.packages`
should NOT appear in RMarkdown or R script files because then they are 
needlessly executed every time the commands are executed.

# Coding Style

You should get in the habit of writing R code with good and consistent style. 
Consequently, homework will be marked off for poor and / or inconsistent style.
See Wickham (http://adv-r.had.co.nz/Style.html) for a style guide,
which we will now walk through.

# Data Structures in R

Data in R can take a variety of forms, Wickham's other book includes the following table

Dimensions   | Homogenous       | Heterogeneous
-- | ---------------- | -------------
1d | atomic `vector`  | `list`
2d | `matrix`         | `data.frame`
nd | `array`          |

There are no "scalars" in R but you can have a `vector` of length one, which may or may
not be numeric.

The `str` function is useful for understanding the structure of an object
```{r str}
data("USJudgeRatings")
str(USJudgeRatings)
```
and the same information is available by clicking on the arrow icon to the left of the
object in the Environment tab in the top right.

## The `vector`

A `vector` in R has three important components

1. Its type, which is given by the `typeof` function
2. Its length, which is given by the `length` function
3. Its attributes, which are metadata given by the `attributes` function

A vector is usually one of four types

1. numeric, which is to say double-precision real
2. integer, which is a signed integer
3. character, which is anything in quotation marks
4. logical, which is `TRUE` or `FALSE`

We can use the `c()` function to create a `vector`
```{r}
num <- c(3.5, 10)           # intermediate flexibility
int <- c(-2L, 4L)           # intermediate flexibility
char <- c("hello", "world") # most flexible
logic <- c(TRUE, FALSE)     # least flexible
```
But any vector can also have `NA` to indicate a missing value. Each of the four
main types of vectors can be tested with the corresponding `is.*` function
```{r test_vector}
is.numeric(num)
is.integer(int)
is.character(char)
is.logical(logic)
is.character(logic)
```
There are also `as.*` functions that can be used to coerce a vector to another
type of vector
```{r coercion}
as.numeric(int)
as.integer(num)
as.character(num)
as.logical(char)
```

### The `factor`

A `factor` is essentially a special kind of vector that is used to represent
categorical data. R is much stricter about categorical data than most other 
statistical programs, which many people find pedantic but only because they 
have not thought about such things carefully. A factor defines a correspondence
between integer codes and their descriptive labels and is usually created by
the `factor` function, such as
```{r}
religion <- factor(c(1, 1, 2, 1, 3, 1, 1, 1, 4), levels = 1 : 4, 
                  labels = c("Christian", "Jewish", "Muslim", "Other"))
levels(religion)
```
This can be used if you have data on individuals' religions where there are
four possible values. We tend to think of religions as being unordered, but
some factors have an unambiguous order such as
```{r}
approval <- factor(c(1, 3, 2, 2, 3, 1), levels = 1 : 3, 
                   labels = c("Disapprove", "Neutral", "Approve"), ordered = TRUE)
nlevels(approval)
```
Be aware that if you do not specify the `levels` and `labels` arguments, R will
use alpha-numerical ordering of the categories, regardless of whether `ordered = TRUE`
and alpha-numerical categories are often arbitrary.

## The `matrix` and the `array`

A `matrix` is a collection of vectors that have the same length. Again, you
can have a matrix whose elements are characters, in which case linear algebra
operations will not work. The usual way to create a matrix is with the `matrix()`
function, such as
```{r}
mat <- matrix(1 : 6, nrow = 3, ncol = 2)
mat
```
There are three important functions for checking the dimensionality of a `matrix`
```{r}
dim(mat)
nrow(mat)
ncol(mat)
is.matrix(mat)
```
You can also create a matrix by binding vectors, either column-wise or row-wise
using the `cbind()` and `rbind()` functions respectively, as in
```{r}
mat <- cbind(6 : 4, 3 : 1)
mat
mat <- rbind(1 : 2, 3 : 4, 5 : 6)
mat
```
This last matrix could also be created simply by specifying `byrow = TRUE`
in the call to the `matrix()` function:
```{r}
mat <- matrix(1 : 6, nrow = 3, ncol = 2, byrow = TRUE)
mat
```
An `array` is essentially a multidimensional generalization of vectors
and matrices (but without linear algebra operations). Conversely, a vector is 
a one-dimensional array and a matrix is a two-dimensional array. Arrays are 
typically created with the `array()` function, as in
```{r}
arr <- array(1 : 24, dim = c(4, 2, 3)) # 4 rows, 2 columns, 3 shelves
arr
dim(arr)
nrow(arr)
ncol(arr)
is.matrix(arr)
is.array(arr)
```

## The `list`

A `list` is a container of containers and its elements need not be of the
same type. Thus, this is a list containing a vector, a factor, a matrix, and 
an array that we created above
```{r}
lst <- list(logic, approval, mat, arr)
is.list(lst)
length(lst)
```
A `list` can contain another `list` as one of its elements. A `list` can have
named elements, although this one currently does not
```{r}
names(lst)
```
So, we can assign the names as follows
```{r}
names(lst) <- c("first", "second", "third", "fourth")
```
but we could use any character vector of length $4$ for the names. Many R functions
return a `list` in order to hold heterogenous information.

## The `data.frame`

A `data.frame` is a special case of a `list` where every list element has the
same length. Thus, a `data.frame` resembles a matrix conceptually but unlike
a `matrix` in R, the "columns" of a `data.frame` need not be the same type.
Here is a `data.frame` of heterogenous vectors
```{r}
df <- data.frame(num, char, logic)
is.data.frame(df)
dim(df) # 2 rows, 3 columns
df
```
Thus, a `data.frame` can also accomodate a mix of factors and numeric data,
which is customary for datasets in the social sciences. When you read data into
R, the output of the function will usually be a `data.frame`.

You can inspect the column (or row names) of a `data.frame` with
```{r}
colnames(df)
rownames(df)
```
and these can have a character vector assigned to them, such as
```{r}
rownames(df) <- c("one", "two")
rownames(df)
```
You can coerce a `matrix` into a `data.frame` (and vice versa)
```{r, eval=FALSE}
df <- as.data.frame(mat)
```
Like a `matrix`, it is possible to use the `cbind` or `rbind` function to
bind two `data.frame`s together by column or by row respectively.

# The `tibble`

Wickham has a new R package called __tibble__ that introduces a new type
of object, also called a `tibble` that behaves pretty much like a 
`data.frame` but avoids some of the ideosyncracies of `data.frame`s.

A `tibble`, which is defined in the _tibble_ package is mostly compatible with a 
`data.frame` but eliminates some characteristics of `data.frame`s that Wickham
considers outdated and / or quirky. One can be created by calling `tibble()` or
equivalently `data_frame` after having loaded the **tibble** package or a package
that imports the **tibble** package:
```{r}
library(tibble)
as_tibble(iris)
my_tibble <- tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

A `tribble` is essentially a transposed `tibble` that is perhaps useful for
manual data entry.

# Functions for Reading and Writing data

R can read data from and write data to a variety of formats. The simplest format
is delimited text, whether locally on your hard-disk or from a URL. However, R
by default cannot read data from password-protected URLs so it is necessary in those
cases to first download the data to your local hard-disk.

For example, you can write a `data.frame` to the working directory on your hard disk
by executing
```{r, message = FALSE}
library(readr)
write_csv(USJudgeRatings, path = "USJudgeRatings.csv")
```
which will save it in so-called "comma-delimited" format.

RStudio has a dedicated mechanism for reading delimited text that can be accessed
by clicking __File -> Import Dataset__. There will be several options in the
wizard, which I will show now

When everything is set, click the __Import__ button and the resulting `tibble` 
will be created in R's workspace. RStudio will also call its `View()` function on 
that `tibble` by default. At this point, it is _very_ important that you look in the 
__Console__ panel of RStudio to see the command that has been piped in. Copy and
paste this line into your .R or .Rmd file so that when you `source` or `knit` it, 
your data will be imported programatically rather than you having to manually go 
through the wizard again. But _do_ edit the line so that it refers to a _relative_ 
path rather than an absolute path. This line could involve the `readr::read_csv` 
function, which is the underlying functions behind the wizard and 
offer some additional arguments that are rarely-used but occassionally essential.

Sometimes, you will be faced with a file of data that is not delimited by any 
particular character but rather comes with an index file that says how many
characters each variable contains. This format is called "fixed width" and can
be imported into R with the `read_fwf` function. This function might be called
with something like
```{r, eval=FALSE}
dataset <- read_fwf(file = file.path("relative", "path", "to", "dataset.txt"),
                    col_positions = c(2, 5, 3, 1, 1, 4))
```
which tells R to look on your disk for a file at `C:\path\to\my\datset.txt` on 
Windows or `/path/to/my/dataset.txt` on anything but Windows with six variables of
width 2, 5, 3, 1, 1, and 4 characters respectively that has variable names in
the first row. If successful, a `tibble` called `dataset` (which is poorly
named) will be created in R's workspace.

Objects in R's workspace can be written in a binary format to your hard-disk using 
the `save` function, whose first few arguments are the names of the objects that 
you want to save and also takes a required `file` argument that specifies the relative
path to save them collectively to. By convention, such files end with an .RData 
extension.
```{r}
sewage <- haven::read_dta("http://www.stata-press.com/data/r14/sewage.dta")
save(sewage, df, file = "sewage.RData")
```

This file can be read into R with `load(file = "sewage.RData")`, in which case the
`sewage` and `df` objects will exist in R's workspace even if they were created in a 
different R session (or on a different computer). Clicking __Session->Load Workspace__
allows you to find the .RData file with a GUI and then calls `load`. But remember to
edit what appears in the Console to refer to a relative path after pasting that line
into an R script in the Editor panel.

A better approach to saving particular R objects is to do
```{r}
saveRDS(sewage, file = "sewage.rds")
```
in which case you can subsequently call `readRDS("sewage.rds")` and assign the result
to any symbol you want.

# Subsetting R Objects

If you have one of the containers discussed earlier,
then you can subset them using expressions within brackets.

* By position, with an integer (vector)
    ```{r}
    letters[4]
    letters[c(1,5,9,15,21)]
    ```
* Conversely, you can obtain all elements except those specified with a 
  negative integer (vector)
    ```{r}
    letters[-4]
    letters[c(-2,-5)]
    ```
* With a logical vector of the same length as the object being subsetted
    ```{r}
    char[logic]
    char[!logic]
    ```
* By name, with a character (vector). In the case of a matrix(-like) object,
  you can subset on either the rownames, the columnnames, or both, as in:
    ```{r}
    state.x77["New York", ]
    state.x77[ , c("Population", "Murder")]
    state.x77["New York", c("Population", "Murder")]
    ```
    However, a `tibble` does not have rownames.

In the case of a list(-like) object, there are three ways to subset

1. With "single brackets", the result is a shorter list
    ```{r}
    lst[2:3]
    ```
2. With "double brackets", the result is the corresponding list element
    ```{r}
    lst[[1]]
    ```
3. With the dollar sign, the result is the list element with that name
    ```{r}
    USJudgeRatings$PHYS
    ```
The dollar sign notation is particularly useful in the R Console because you can 
type the first few letters of the element name and then press the Tab key to 
show the possible completions if you do not remember how to spell the element
name exactly.

In all of these cases, it is also possible to assign to a subset of a container.
For example,
```{r}
char[2] <- "Newman"
df$char <- char
df[ ,"char"]
```

To review,

* Parentheses, `()`, are used in two very different situations:
    * Grouping in mathematical expressions, such as
    ```{r}
    (2 + 3) * 4
    ```
    * Function calls, such as
    ```{r}
    log(10)
    ```
* Braces, `{}`, are used in two similar situations:
    * Function definitions
    ```{r}
    sinc <- function(x) {
      return(ifelse(x == 0, 1, sin(x) / x))
    }
    ```
    * Control flow expressions such as `if`, `else`, `for`, `while`, etc.
    ```{r}
    counter <- 1
    while(counter <= 3) {
      print(paste("counter is", counter))
      counter <- counter + 1
    }
    ```
* Brackets, `[]`, are used for subsetting, as in
    ```{r, eval = FALSE}
    x[1]
    mean(x[x > 0])
    ```
An example that illustrates most of these constructions simultaneously is
```{r, eval = FALSE}
len <- length(x)
sinc_values <- rep(NA_real_, len)
for(i in 1:len) {
  if(x[i] == 0) {
    sinc_values[i] <- 1
  }
  else {
    sinc_values[i] <- sin(x[i]) / x[i]
  }
}
```

# Functions

As with everything in R, functions are objects. But functions are arguably the most important
kind of object in a language like R. Non-primitive functions have three important parts

1. The `body`, which executes assignment statements using culminating in a call to `return`
2. The `formals`, which are the arguments to the function
3. The `environment`, which is the location in R's memory where the function is evaluated

For example,
```{r}
f <- function(x) x^2
body(f)
formals(f)
environment(f)
```

It is considered poor form in this class to define functions that rely on 
objects that are neither passed to the function as arguments or created inside the
function. In other words, a function typically should not utilize objects in its
environment.

Function arguments are matched

1. by exact name
2. by partial name
3. by position

in that order. However, using a partial name is discouraged and matching by position is
subject to subtle errors. Therefore, it is preferable to use exact names, except for
perhaps the first argument to a function. For example,
```{r}
round(x = 2.3, digits = 0) # good
round(2.3, digits = 0) # acceptable
round(2.3, dig = 0) # stupid
round(2.3, 0) # hard to understand
round(2.3) # acceptable because the default value for digits is zero
```
Many R functions have default arguments that are used if the argument is not
specified explicitly. It is generally good practice to not specify default
arguments if you do not want to change their values. Consequently, when
you see an argument specified, it is clear that the programmer's intent
is to depart from the default value
```{r}
round(2.34)
round(2.34, digits = 1)
```
You can use the `do.call` function to call a function by passing in a 
list of its arguments, such as
```{r}
do.call(mean, args = list(x = rnorm(10), trim = 0.1))
```
Finally, many R functions have `...` in their arguments, such as
```{r}
args(mean)
args(mean.default)
```
This means that any additional arguments that are passed to the `mean`
function besides `x` are passed along to any function that the `mean`
function might subsequently call. For example, if `mean` receives a
numeric `vector` for its `x` argument, the `mean` function will call the
`mean.default` function that has two additional arguments `trim` (with
a default of `0`) and `na.rm` (with a default of `FALSE`). This implies,
by default, that R will calculate an untrimmed mean and will not remove
any missing values (which will cause `mean` to return `NA`). But if we
pass `na.rm = TRUE` to the `mean` function, it will capture that argument
in its `...` and pass that along to `mean.default`, which will then calculate
the sample mean excluding any `NA` values. In code,
```{r}
vals <- c(2, 5, NA)
mean(vals) # NA
mean(vals, na.rm = TRUE)
```

The part about R that students find the most confusing is that an R function
function with named arguments just uses those names internally and for
documentation. They have essentially nothing to do with what names you have used
to bind results to. For example, you might assume that the following would work,
since `x` is the name of the first argument of the `round` function:
```{r, eval=FALSE}
x <- 5.1
round() # yields an error
```
but in fact, it produces an error message that in essence says that R does not
know what you want it to round. In contrast, this does work
```{r}
x <- 5.1
round(x = 7/2) # x is merely the internal symbol that 7/2 is bound to
x # x is still bound to 5 outside the round() function
round(x) # specifing by position
round(x = x) # specifying by name
```
This is why it is considered unacceptable in G5058 to use the "single equals"
sign for assignment, even though it is technically legal in R and obligatory
in many other languages like C++. In R, `<-` should be used to assign a symbol
to something and `=` should be used to specify function arguments by name.

# Swirl

The __swirl__ package is good for learning R. It asks you questions and gives you a prompt
where you can type your answer.

```{r, eval=FALSE}
library(swirl)
swirl()
```

